---
title: "Biostat 203B Homework 1"
author: Kuan-Hung Yeh (UID:705669455)
subtitle: Due Jan 21 @ 11:59PM
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Display machine information for reproducibility:
```{r, echo=TRUE}
sessionInfo()
```

## Q1. Git/GitHub

**No handwritten homework reports are accepted for this course.** We work with Git and GitHub. Efficient and abundant use of Git, e.g., frequent and well-documented commits, is an important criterion for grading your homework.

1. Apply for the [Student Developer Pack](https://education.github.com/pack) at GitHub using your UCLA email. You'll get GitHub Pro account for free (unlimited public and private repositories).

2. Create a **private** repository `biostat-203b-2022-winter` and add `Hua-Zhou` and `maschepps` as your collaborators with write permission.

3. Top directories of the repository should be `hw1`, `hw2`, ... Maintain two branches `main` and `develop`. The `develop` branch will be your main playground, the place where you develop solution (code) to homework problems and write up report. The `main` branch will be your presentation area. Submit your homework files (R markdown file `Rmd`, `html` file converted from R markdown, all code and extra data sets to reproduce results) in `main` branch.

4. After each homework due date, teaching assistant and instructor will check out your main branch for grading. Tag each of your homework submissions with tag names `hw1`, `hw2`, ... Tagging time will be used as your submission time. That means if you tag your `hw1` submission after deadline, penalty points will be deducted for late submission.

5. After this course, you can make this repository public and use it to demonstrate your skill sets on job market.

## Q2. Data ethics training

This exercise (and later in this course) uses the [MIMIC-IV data](https://mimic-iv.mit.edu), a freely accessible critical care database developed by the MIT Lab for Computational Physiology. Follow the instructions at <https://mimic.mit.edu/docs/gettingstarted/> to (1) complete the CITI `Data or Specimens Only Research` course and (2) obtain the PhysioNet credential for using the MIMIC-IV data. Display the verification links to your completion report and completion certificate here. (Hint: The CITI training takes a couple hours and the PhysioNet credentialing takes a couple days; do not leave it to the last minute.)

Verification links:
<br>
completion report <www.citiprogram.org/verify/?ka3709a6c-2c5e-48e2-96f4-b5d4b5d73122-46525347>
<br>
completion certificate <www.citiprogram.org/verify/?w5b2ce486-632e-4483-8bc5-7ef810042545-46525347>
<br>
PhysioNet credential <https://drive.google.com/file/d/1yIJfb0cJrOzLICmkzgp5q3jr3Whr8GK9/view?usp=sharing>

## Q3. Linux Shell Commands

1. The `/mnt/mimiciv/1.0` folder on teaching server contains data sets from MIMIC-IV. Refer to the documentation <https://mimic.mit.edu/docs/iv/> for details of data files.  
```{bash, echo=T}
ls -l /mnt/mimiciv/1.0
```
Please, do **not** put these data files into Git; they are big. Do **not** copy them into your directory. Do **not** decompress the gz data files. These create unnecessary big files on storage and are not big data friendly practices. Just read from the data folder `/mnt/mimiciv/1.0` directly in following exercises. 

    Use Bash commands to answer following questions.

2. Display the contents in the folders `core`, `hosp`, `icu`. Why are these data files distributed as `.csv.gz` files instead of `.csv` (comma separated values) files? Read the page <https://mimic.mit.edu/docs/iv/> to understand what's in each folder.
```{bash, echo=T}
ls -l /mnt/mimiciv/1.0/core
```
```{bash, echo=T}
ls -l /mnt/mimiciv/1.0/hosp
```
```{bash, echo=T}
ls -l /mnt/mimiciv/1.0/icu
```

As we know that csv.gz files are the CSV file compressed with gzip utility. 
The reason why those data distributed as .csv.gz files are because they are too 
large. In order to reduce the size of files, they used compressed file type csv.gz

3. Briefly describe what bash commands `zcat`, `zless`, `zmore`, and `zgrep` do.

#man zcat
'The zcat command allows the user to expand and view a compressed file without 
uncompressing that file.'
<br>
#man zless
'Zless is a filter which allows examination of compressed or plain text files 
one screenful at a time on a soft-copy terminal.'
<br>
#man zmore
'Zmore command is a handy utility that lets you view gzipped files one screen 
at a time (much like the more command) without first decompressing them.'
<br>
#man zgrep
'Zgrep command is used to search out expressions from a given a file even if 
it is compressed.'

4. What's the output of following bash script?
Listing all the csv.gz files contents in /mnt/mimiciv/1.0/core/*.gz directory. 
```{bash, echo=T}
    for datafile in /mnt/mimiciv/1.0/core/*.gz
      do
        ls -l $datafile
      done
```
Display the number of lines in each data file using a similar loop.

Answer: the outcomes for this command are included the header line.

```{bash, echo=T}
    for datafile in /mnt/mimiciv/1.0/core/*.gz
      do
       zcat $datafile | wc -l
      done
```


5. Display the first few lines of `admissions.csv.gz`. How many rows are in this data file? How many unique patients (identified by `subject_id`) are in this data file? (Hint: combine Linux commands `zcat`, `head`/`tail`, `awk`, `sort`, `uniq`, `wc`, and so on.)
```{bash, echo=T}
zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | head
#Display the first few lines of `admissions.csv.gz`.
```
```{bash, echo=T}
zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | tail -n +2 | wc -l
#There are 523740 rows (after remove header) in this data file.
```
```{bash, echo=T}
zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | tail -n +2 | awk -F, '{print $1}' | sort |
uniq | wc -l
#There are 256878 (after remove header) unique patients in this data file
```
6. What are the possible values taken by each of the variable `admission_type`, `admission_location`, `insurance`, and `ethnicity`? Also report the count for each unique value of these variables. (Hint: combine Linux commands `zcat`, `head`/`tail`, `awk`, `uniq -c`, `wc`, and so on.)
```{bash, echo=T}
#admission_type
zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | tail -n +2 | awk -F, '{print $6}' | sort |
uniq -c | wc -l
zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | tail -n +2 | awk -F, '{print $6}' | sort |
uniq -c | head -10
#There are 9 unique admission_type value in this data file
```
```{bash, echo=T}
#admission_location
zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | tail -n +2 | awk -F, '{print $7}' | sort |
uniq -c | wc -l
zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | tail -n +2 | awk -F, '{print $7}' | sort |
uniq -c | head -13
#There are 12 unique admission_location value in this data file
```
```{bash, echo=T}
#insurance
zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | tail -n +2 | awk -F, '{print $9}' | sort |
uniq -c | wc -l
zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | tail -n +2 | awk -F, '{print $9}' | sort |
uniq -c | head
#There are 3 unique insurance value in this data file
```
```{bash, echo=T}
#ethnicity
zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | tail -n +2 | awk -F, '{print $12}' | sort |
uniq -c | wc -l
zcat /mnt/mimiciv/1.0/core/admissions.csv.gz | tail -n +2 | awk -F, '{print $12}' | sort |
uniq -c | head
#There are 8 unique ethnicity value in this data file
```

## Q4. Who's popular in Price and Prejudice

1. You and your friend just have finished reading *Pride and Prejudice* by Jane Austen. Among the four main characters in the book, Elizabeth, Jane, Lydia, and Darcy, your friend thinks that Darcy was the most mentioned. You, however, are certain it was Elizabeth. Obtain the full text of the novel from <http://www.gutenberg.org/cache/epub/42671/pg42671.txt> and save to your local folder. 
```{bash, echo=T}
wget -nc http://www.gutenberg.org/cache/epub/42671/pg42671.txt
```
Explain what `wget -nc` does. Do **not** put this text file `pg42671.txt` in Git. Complete the following loop to tabulate the number of times each of the four characters is mentioned using Linux commands.

Wget is the non-interactive network downloader which is used to download 
files from the server even when the user has not logged on to the system 
and it can work in the background without hindering the current process.
```{bash, echo=T}
wget -nc http://www.gutenberg.org/cache/epub/42671/pg42671.txt
for char in Elizabeth Jane Lydia Darcy
do
  echo $char:
  grep $char pg42671.txt | wc -l
done
```

2. What's the difference between the following two commands?
```{bash eval=FALSE}
echo 'hello, world' > test1.txt
```
    and
```{bash eval=FALSE}
echo 'hello, world' >> test2.txt
```
For the first command, it directs 'hello, world' from command to test1.txt.
However, for the second, it appends 'hello, world' from command to test1.txt.

3. Using your favorite text editor (e.g., `vi`), type the following and save the file as `middle.sh`:
```{bash eval=FALSE}
#!/bin/sh
# Select lines from the middle of a file.
# Usage: bash middle.sh filename end_line num_lines
head -n "$2" "$1" | tail -n "$3"
```

```{bash, echo=T}
#vi ./middle.sh 
# then copy and paste the content
```
Using `chmod` make the file executable by the owner, and run 
```{bash, echo=T}
chmod 751 /home/khyeh0816/biostat-203b-2022-winter/middle.sh
ls -al
#-rwxr-x--x.  1 khyeh0816 khyeh0816  124 Jan 14 23:42 middle.sh
```

```{bash, echo=T}
/home/khyeh0816/biostat-203b-2022-winter/middle.sh pg42671.txt 20 5
#head -n 20 pg42671.txt | tail -n 5
```
Explain the output. Explain the meaning of `"$1"`, `"$2"`, and `"$3"` in this shell script. Why do we need the first line of the shell script?

The output for this command prints author and editor for *Pride and Prejudice*.
For the `"$1"`, `"$2"`, and `"$3"` in this shell script, `"$1"` means that the
filename of this command (In this scenario is *pg42671.txt*). For the `"$2"`, it
means that the end_line for this file which is number 20th rows in this case.
For the `"$3"`, it shows that how many lines we wants in this case is 5 lines.
Thus, the command above is just same as I wrote below:
```{bash, echo=T}
head -n 20 pg42671.txt | tail -n 5
```
Lastly, the reason why we need first line of the shell script (#!/bin/sh) is 
because we would like to make the script always be run with sh,
rather than another shell.

### Q5. More fun with Linux

Try following commands in Bash and interpret the results: `cal`, `cal 2021`, `cal 9 1752` (anything unusual?), `date`, `hostname`, `arch`, `uname -a`, `uptime`, `who am i`, `who`, `w`, `id`, `last | head`, `echo {con,pre}{sent,fer}{s,ed}`, `time sleep 5`, `history | tail`.
```{bash, echo=T}
cal
```
For this result, cal means that printing the calender for this month (Jan, 2022)

```{bash, echo=T}
cal 2021
```
For this result, cal 2021 means that printing the calender for last whole year
(2021).

```{bash, echo=T}
cal 9 1752
```
For this result, cal 9 1752 means that printing the calender for September 1752.
The unusual thing is that the date 3rd to 13th are missing for this calender.

```{bash, echo=T}
date
```
For this result, date means that printing the date and time right now.

```{bash, echo=T}
hostname
```
For this result, hostname means that printing DNS(Domain Name System) name now.

```{bash, echo=T}
arch
```
For this result, arch means that printing the computer architecture now.

```{bash, echo=T}
uname -a
```
For this result, uname -a means that printing System information used right now.

```{bash, echo=T}
uptime
```
For this result, uptime returns information about how long your system has been 
running together with the current time, number of users with running sessions, 
and the system load averages for the past 1, 5, and 15 minutes. 

```{bash, echo=T}
whoami
```
For this result, whoami returns information about current username.

```{bash, echo=T}
who
```
For this result, who command lets you display the users currently logged in to 
your UNIX or Linux operating system.

```{bash, echo=T}
w
```
For this result, w commmand used to show who is logged on and 
what they are doing.

```{bash, echo=T}
id
```
For this result, id commmand used to show your current ID for this system.

```{bash, echo=T}
last | head
```
For this result, last | head used to display top 10 (head) the list of all the 
users logged in and out.

```{bash, echo=T}
echo {con,pre}{sent,fer}{s,ed}
```
For this result, echo command used to display line of text/string that are 
passed as an argument. In this case, it shows the permutations for the words 
inside those three braces.

```{bash, echo=T}
time sleep 5
```
For this result, time sleep command used to delay for a fixed amount of time 
during the execution of any script.

```{bash, echo=T}
history | tail
```

The history command shows a list of the commands entered since you started 
the session.








