---
title: "Biostat 203B Homework 2"
subtitle: Due Feb 6 @ 11:59PM
author: Kuan-Hung (Peter) Yeh (UID:705669455)
output: 
  html_document:
    toc: true
    toc_depth: 4 
---

Display machine information for reproducibility:
```{r, echo=TRUE}
sessionInfo()
```

```{r setup, message=F}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, cache.lazy = FALSE)
library(tidyverse)
library(data.table)
library(lubridate)
```

```{r}
os <- sessionInfo()$running
if (str_detect(os, "Linux")) {
  mimic_path <- "/mnt/mimiciv/1.0"
} else if (str_detect(os, "macOS")) {
  mimic_path <- "/Users/huazhou/Documents/Box Sync/MIMIC/mimic-iv-1.0"
}
```

In this exercise, we use tidyverse (ggpot2, dplyr, etc) to explore the [MIMIC-IV](https://mimic.mit.edu/docs/iv/) data introduced in [homework 1](https://ucla-biostat-203b.github.io/2022winter/hw/hw1/hw1.html) and to build a cohort of ICU stays.

```{r}
# tree -s -L 2 /Users/huazhou/Documents/Box\ Sync/MIMIC/mimic-iv-1.0
system(str_c("tree -s -L 2 ", shQuote(mimic_path)), intern = TRUE)
```

## Q1. `read.csv` (base R) vs `read_csv` (tidyverse) vs `fread` (data.table)

There are quite a few utilities in R for reading plain text data files. Let us test the speed of reading a moderate sized compressed csv file, `admissions.csv.gz`, by three programs: `read.csv` in base R, `read_csv` in tidyverse, and `fread` in the popular data.table package. 

```{r, echo=TRUE}
#`read.csv` in base R
system.time(read.csv("/mnt/mimiciv/1.0/core/admissions.csv.gz"))
#`read_csv` in tidyverse
system.time(read_csv("/mnt/mimiciv/1.0/core/admissions.csv.gz"))
# `fread` in the popular data.table package.
system.time(fread("/mnt/mimiciv/1.0/core/admissions.csv.gz"))
```

Which function is fastest? Is there difference in the (default) parsed data types? (Hint: R function `system.time` measures run times.)

**Solution:**

```{r, echo=TRUE}
#`read.csv` in base R
read1 <- read.csv("/mnt/mimiciv/1.0/core/admissions.csv.gz")
str(read1)
#`read_csv` in tidyverse
read2 <- read_csv("/mnt/mimiciv/1.0/core/admissions.csv.gz")
str(read2)
# `fread` in the popular data.table package.
read3 <- fread("/mnt/mimiciv/1.0/core/admissions.csv.gz")
str(read3)
```

The fread in the data.table package is the fastest function to read a moderate 
sized compressed csv file. From the data type table above, we can see that the
default data type for read_csv and fread are the same. However, the default data
type for read.csv is differ from the other two function. The data structure all 
are data.frame, but the class of variable are different.

For later questions, we stick to the tidyverse.

## Q2. ICU stays

`icustays.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/icustays/>) contains data about Intensive Care Units (ICU) stays. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/icu/icustays.csv.gz")), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Import `icustatys.csv.gz` as a tibble `icustays_tble`. 

**Solution:**

```{r, echo=TRUE}
#`read_csv` in tidyverse to import icustatys.csv.gz
icustays_tble <- read_csv("/mnt/mimiciv/1.0/icu/icustays.csv.gz")
head(icustays_tble)
```
2. How many unique `subject_id`? Can a `subject_id` have multiple ICU stays? 

**Solution:**

```{r, echo=TRUE}
icustays_tble %>%
distinct(subject_id) %>%
nrow()
```

There are 53150 unique subject_id in icustays_tble. And, Yes, a subject id can 
have more than one ICU stays. From the result below, we can see that the maximum 
time of ICU stays is 37 times for subject is = 18358138.

```{r, echo=TRUE}
icustays_tble %>%
group_by(subject_id) %>%
summarise(n=n()) %>%
arrange(desc(n)) %>%
head()
```
3. For each `subject_id`, let's only keep the first ICU stay in the tibble `icustays_tble`.

**Solution:**

From the result below, I checked the rows of this subset will be 53150, which 
make sense sine every nuique patient had their first record when they 
transferred to ICU.

```{r, echo=TRUE}
icustays_tble_1st_visit <- icustays_tble %>%
                  group_by(subject_id) %>%
                  arrange(intime) %>%
                  filter(row_number()==1)
icustays_tble_1st_visit
```
## Q3. `admission` data

Information of the patients admitted into hospital is available in `admissions.csv.gz`. See <https://mimic.mit.edu/docs/iv/modules/core/admissions/> for details of each field in this file. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/core/admissions.csv.gz")), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Import `admissions.csv.gz` as a tibble `admissions_tble`.

**Solution:**

```{r, echo=TRUE}
admissions_tble <- read_csv("/mnt/mimiciv/1.0/core/admissions.csv.gz")
head(admissions_tble)
```
2. Let's only keep the admissions that have a match in `icustays_tble` according to `subject_id` and `hadmi_id`.

**Solution:**

```{r, echo=TRUE}
ad_icu_combine <-
  admissions_tble %>% 
  semi_join(icustays_tble, by = c("subject_id", "hadm_id"))
```

3. Summarize the following variables by graphics. 

**Solution:**

- admission year  

```{r, echo=TRUE}
#admission year 
ad_icu_combine$admittime <- ymd_hms(ad_icu_combine$admittime)
ad_icu_combine %>% 
  mutate(year = year(admittime), label = TRUE) %>% 
  ggplot(aes(x = year)) +
  geom_bar()
```
- admission month  

```{r, echo=TRUE}
#admission month
ad_icu_combine %>% 
  mutate(month = month(admittime), label = TRUE) %>% 
  ggplot(aes(x = month)) +
  geom_bar()
```
- admission month day  

```{r, echo=TRUE}
#admission month day
ad_icu_combine %>% 
  mutate(mday = mday(admittime), label = TRUE) %>% 
  ggplot(aes(x = mday)) +
  geom_bar()
```
- admission week day  

```{r, echo=TRUE}
#admission week day
ad_icu_combine %>% 
  mutate(wday = wday(admittime), label = TRUE) %>% 
  ggplot(aes(x = wday)) +
  geom_bar()
```
- admission hour (anything unusual?)  

```{r, echo=TRUE}
#admission hour
ad_icu_combine %>% 
  mutate(hour = hour(admittime), label = TRUE) %>% 
  ggplot(aes(x = hour)) +
  geom_bar()
```

From the plot above, we can see the trend that the most frequent hour in a day 
that people got transferred to ICU is during the midnight (00:00 AM-01:00 AM). 
Moreover, the ICU cases from 15:00 PM to 00:00 AM tend to be higher than the 
cases number from 01:00 AM to 15:00 PM.

## Q4. `patients` data

Patient information is available in `patients.csv.gz`. See <https://mimic.mit.edu/docs/iv/modules/core/patients/> for details of each field in this file. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/core/patients.csv.gz")), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Import `patients.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/core/patients/>) as a tibble `patients_tble` and only keep the patients who have a match in `icustays_tble` (according to `subject_id`).

**Solution:**

```{r, echo=TRUE}
patients_tble <- read_csv("/mnt/mimiciv/1.0/core/patients.csv.gz")
patient_icu_combine <-
  patients_tble %>% 
  inner_join(icustays_tble, by = "subject_id")
patient_icu_combine
```

2. Summarize variables `gender` and `anchor_age`, and explain any patterns you see.

**Solution:**

```{r, echo=TRUE}
#gender
patient_icu_combine %>%
  ggplot(mapping = aes(x = gender, fill = gender)) + 
  geom_bar(width = 1)
```

From the plot above, we can tell that the number of male is larger than female.

```{r, echo=TRUE}
#anchor_age
patient_icu_combine %>%
  ggplot(mapping = aes(x = anchor_age, fill = anchor_age)) + 
  geom_bar(width = 1)
summary(patient_icu_combine$anchor_age)
```

From the plot above, we can tell that the maximum age for the patient is 91, which
is also the most frequent age in this data. That's weird because it is not possible that
the number of most frequent age appear in the extreme value. I believe that there must be
an age transformation for anchor_age. On the other hand, the distribution of the anchor age 
is left skewed with the median anchor age is 65.


## Q5. Lab results

`labevents.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/hosp/labevents/>) contains all laboratory measurements for patients. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/hosp/labevents.csv.gz")), 
    " | head"
    ), 
  intern = TRUE
)
```
`d_labitems.csv.gz` is the dictionary of lab measurements. 
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/hosp/d_labitems.csv.gz")), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Find how many rows are in `labevents.csv.gz`.

2. We are interested in the lab measurements of creatinine (50912), potassium (50971), sodium (50983), chloride (50902), bicarbonate (50882), hematocrit (51221), white blood cell count (51301), glucose (50931), magnesium (50960), and calcium (50893). Retrieve a subset of `labevents.csv.gz` only containing these items for the patients in `icustays_tble` as a tibble `labevents_tble`. 

    Hint: `labevents.csv.gz` is a data file too big to be read in by the `read_csv` function in its default setting. Utilize the `col_select` and `lazy` options in the `read_csv` function to reduce the memory burden.

3. Further restrict `labevents_tble` to the first lab measurement during the ICU stay. 
4. Summarize the lab measurements by appropriate numerics and graphics. 

## Q6. Vitals from charted events

`chartevents.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/chartevents/>) contains all the charted data available for a patient. During their ICU stay, the primary repository of a patient’s information is their electronic chart. The `itemid` variable indicates a single measurement type in the database. The `value` variable is the value measured for `itemid`. The first 10 lines of `chartevents.csv.gz` are
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/icu/chartevents.csv.gz")), 
    " | head"), 
  intern = TRUE
)
```
`d_items.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/d_items/>) is the dictionary for the `itemid` in `chartevents.csv.gz`. 
```{r}
system(
  str_c(
    "zcat < ", 
    shQuote(str_c(mimic_path, "/icu/d_items.csv.gz")), 
    " | head"), 
  intern = TRUE
)
```

1. We are interested in the vitals for ICU patients: heart rate (220045), mean non-invasive blood pressure (220181), systolic non-invasive blood pressure (220179), body temperature in Fahrenheit (223761), and respiratory rate (220210). Retrieve a subset of `chartevents.csv.gz` only containing these items for the patients in `icustays_tble` as a tibble `chartevents_tble`.

2. Further restrict `chartevents_tble` to the first vital measurement during the ICU stay. 

3. Summarize these vital measurements by appropriate numerics and graphics. 

## Q7. Putting things together

Let us create a tibble `mimic_icu_cohort` for all ICU stays, where rows are  

- first ICU stay of each unique adult (age at admission > 18)

and columns contain at least following variables  

- all variables in `icustays.csv.gz`  
- all variables in `admission.csv.gz`  
- all variables in `patients.csv.gz`  
- first lab measurements during ICU stay  
- first vital measurements during ICU stay
- an indicator variable `thirty_day_mort` whether the patient died within 30 days of hospital admission (30 day mortality)

## Q8. Exploratory data analysis (EDA)

Summarize following information using appropriate numerics or graphs.

- `thirty_day_mort` vs demographic variables (ethnicity, language, insurance, marital_status, gender, age at hospital admission)

- `thirty_day_mort` vs first lab measurements

- `thirty_day_mort` vs first vital measurements

- `thirty_day_mort` vs first ICU unit

